Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-01-25 xrb2/paper.tex
Updated plot of temp/enucdot maxima

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We continue to investigate laterally propagating flames in type I X-ray bursts using
fully compressible hydrodynamics simulations.  In the current study we
relax the approximations used previously and explore the effect of
rotation rate and the thermal structure of the neutron star on the
flame.  We find that at lower rotation rates the flame becomes harder to ignite,
whereas at higher rotation rates the nuclear burning is enhanced and the flame propagates steadily.
We also find that higher temperature atmospheres lead to faster
flames and observe acceleration of the flame front as it propagates
through the atmosphere.  All of the software used for these
simulations is freely available.

@@

We investigate the effect of rotation rate on the flame. With the
exception of IGR J17480-2446 (\citet{altamirano2010discovery}, spinning at $11~\mathrm{Hz}$), most observations of
XRBs come from sources with rotation rates of $200-600~\mathrm{Hz}$
\citep{bilous:2019,galloway:2020}. There are a number of
factors that could explain this lack of observations below $200~\mathrm{Hz}$. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that burst
oscillations at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it does not have
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below $200~\mathrm{Hz}$ 
\MarginPar{look into this?}.

@@

For some of our analysis, we would like to have a means of estimating the temperature ($T$) and nuclear 
energy generation rate ($\enucdot$) in the burning region of each simulation. For this purpose, we define 
the mass weighted average $\repavg{Q}$ of some quantity $Q$ to be
\begin{equation}
	\label{eqn:top_percentile_avg}
	\repavg{Q} \equiv \frac{\sum_{c_i} m(c_i) Q(c_i)}{\sum_{c_i} m(c_i)}; ~c_i \in C_{99}(Q).
\end{equation}
\noindent Here, $C_{99}(Q)$ is the set of grid cells with $Q$ values in the top percentile, and 
$Q(c_i)$ is the value of $Q$ in cell $c_i$ (with $m(c_i)$ being the total mass in cell $c_i$). By 
using $\repavg{Q}$ instead of simply taking the maximum of the quantity across the entire 
simulation domain allows us to track changes over the domain as a whole rather than at a single 
localized point. This will therefore be a better reflection of the overall behavior of the flame 
rather than of a single localized fluctuation. Figure \ref{fig:maxima} shows $\repavg{T}$ and 
$\repavg{\enucdot}$ as functions of time for the subset of our runs that achieve a propagating 
flame. This figure is referenced throughout the subsequent sections.

@@

We ran a number of simulations of laterally propagating flames in XRBs in order to explore the effects of rotation and thermal structure. We found that increasing the rotation rate increased the energy generation rate within the flame and enhanced nuclear burning. Apart from the lowest rotation run (which failed to ignite), flame propagation was not noticeably impacted by rotation rate; by the time the different flame fronts reached steady propagation, they shared comparable velocities. These results are likely due to the rotation-dependent strength of the Coriolis force its confinement of the flame balancing the enhanced nuclear burning. We explored several models with different crustal temperatures to determine what effects mechanisms such as higher accretion rates, crustal heating and shallow heating may have on flame propagation. We found that increasing the temperature of the crust significantly enhanced the flame propagation. This we believe to be because a cooler crust allows heat to more efficiently be transferred away from the flame itself, therefore reducing the flame's temperature, slowing burning and consequently reducing its propagation speed. At higher temperatures, we saw that the flame itself accelerated, the increased temperature and inability for heat to be transported away from the flame front driving unstable, accelerating flame propagation. We saw that if the crust temperature was too high, then instead of a flame the entire atmosphere would burn stably. This is reminiscent of what is seen for neutron stars with accretion rates exceeding the Eddington luminosity.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We continue to investigate two-dimensional laterally propagating flames in type I X-ray bursts using fully compressible hydrodynamics simulations.  In the current study we relax previous approximations where we artificially boosted the flames. We now use more physically realistic reaction rates, thermal conductivities, and rotation rates, exploring the effects of neutron star rotation rate and thermal structure on the flame.  We find that at lower rotation rates the flame becomes harder to ignite, whereas at higher rotation rates the nuclear burning is enhanced by increased confinement from the Coriolis force and the flame propagates steadily. At higher crustal temperatures, the flame moves more quickly and accelerates as it propagates through the atmosphere. If the temperature is too high, , instead of a flame propagating across the surface the entire atmosphere burns steadily. All of the software used for these simulations is freely available.

@@

We investigate the effect of rotation rate on the flame. With the
exception of IGR J17480-2446 (\citealt{altamirano2010discovery}, spinning at $11~\mathrm{Hz}$), most observations of
XRBs come from sources with rotation rates of $200-600~\mathrm{Hz}$
\citep{bilous:2019,galloway:2020}. There are a number of
factors that could explain this lack of observations below $200~\mathrm{Hz}$. It
could be that there is some physical process which inhibits the flame
ignition and/or spread at lower rotation rates. It could be that burst
oscillations at lower rotation rates are smaller in amplitude and
therefore more difficult to detect. It could be that it does not have
anything to do with the flame at all, but that neutron stars in
accreting low mass X-ray binaries rarely have rotation rates below $200~\mathrm{Hz}$ 
\MarginPar{look into this?}.

@@

For some of our analysis, we would like to have a means of estimating the temperature ($T$) and 
nuclear energy generation rate ($\enucdot$) in the burning region of each simulation. For this 
purpose, we define the mass-weighted average $\repavg{Q}$ of some quantity $Q$ to be
\begin{equation}
	\label{eqn:top_percentile_avg}
	\repavg{Q} \equiv \frac{\sum_{c_i} m(c_i) Q(c_i)}{\sum_{c_i} m(c_i)}; ~c_i \in C_{99}(Q).
\end{equation}
\noindent Here, $C_{99}(Q)$ is the set of grid cells with $Q$ values in the top percentile, and 
$Q(c_i)$ is the value of $Q$ in cell $c_i$ (with $m(c_i)$ being the total mass in cell $c_i$). By 
using $\repavg{Q}$ instead of simply taking the maximum of the quantity across the entire 
simulation domain allows us to track changes over the domain as a whole rather than at a single 
localized point. This will therefore be a better reflection of the overall behavior of the flame 
rather than of a single localized fluctuation. Figure \ref{fig:maxima} shows $\repavg{T}$ and 
$\repavg{\enucdot}$ as functions of time for the subset of our runs that achieve a propagating 
flame. This figure is referenced throughout the subsequent sections.

@@

We ran a number of simulations of laterally propagating flames in XRBs in order to explore the effects of rotation and thermal structure. We found that increasing the rotation rate increased the energy generation rate within the flame and enhanced nuclear burning. Apart from the lowest rotation run (which failed to ignite), flame propagation was not noticeably impacted by rotation rate; by the time the different flame fronts reached steady propagation, they shared comparable velocities. These results are likely due to the rotation-dependent strength of the Coriolis force its confinement of the flame balancing the enhanced nuclear burning. We explored several models with different crustal temperatures to determine what effects mechanisms such as higher accretion rates, crustal heating and shallow heating may have on flame propagation. We found that increasing the temperature of the crust significantly enhanced the flame propagation. This we believe to be because a cooler crust allows heat to more efficiently be transferred away from the flame itself, therefore reducing the flame's temperature, slowing burning and consequently reducing its propagation speed. At higher temperatures, we saw that the flame itself accelerated, the increased temperature and inability for heat to be transported away from the flame front driving unstable, accelerating flame propagation. We saw that if the crust temperature was too high, then instead of a flame the entire atmosphere would burn steadily. This is reminiscent of what is seen for neutron stars with accretion rates exceeding the Eddington luminosity.
